\documentclass[english]{famnit-thesis}
% \documentclass[english,cover]{famnit-thesis} % Add cover page for print submission.
% \documentclass[english,review]{famnit-thesis} % Review version (anonymised).

\usepackage{ccicons}   % CC license symbols on the license page.
\usepackage{booktabs}  % Professional-quality tables (\toprule, \midrule, \bottomrule).
\usepackage{amsmath}   % Extended math environments.

% ---------------------------------------------------------------------------
% TODO: Replace the placeholder values below with your actual information.
% ---------------------------------------------------------------------------

\title
    {Zaznavanje osi vozil z nevronskimi mrežami za sisteme tehtanja vozil na mostu}
    {Neural Network-Based Axle Detection for Bridge Weigh-In-Motion Systems}
    {Neural Network-Based Axle Detection for B-WIM Systems.}

\author{[FirstName]}{[LastName]}{[LastName] F.}
\studyprogram{Računalništvo in informatika}
\mentor{[prof. dr. Supervisor Name]}{[Supervisor Name, PhD]}
\date{June}{2026}

% Remove the two lines below if you have no co-mentor.
% \comentor{doc. dr. Ime Somentorja}{Ime Somentorja, PhD}
% \workingcomentor{dr. Ime Delovnega Somentorja}{Ime Delovnega Somentorja, PhD}

% ---------------------------------------------------------------------------
% Keywords (Slovene first, then English)
% ---------------------------------------------------------------------------
\keywords
    {zaznavanje osi, tehtanje vozil na mostu, nevronske mreže, TCN, CNN, časovne serije}
    {axle detection, bridge weigh-in-motion, neural networks, temporal convolutional network, time series}

% ---------------------------------------------------------------------------
% Abstracts (Slovene first, then English; max 250 words each)
% ---------------------------------------------------------------------------
\abstract
{
    Sistemi tehtanja vozil na mostu (B-WIM) ocenjujejo bruto težo tovornih vozil
    z analizo dinamičnega signala deformacije mostu med prehodom vozila. Ključni
    korak je natančno zaznavanje osi — določitev točnih vzorčnih pozicij vsakega
    prehoda osi v signalu dolžine 1.300 vzorcev. Gre za zahteven problem dvojiškega
    označevanja zaporedja, ki ga otežuje izrazito neravnovesje razredov: le
    0{,}35\,\% vzorcev je pozitivnih (dogodki osi), kar daje razmerje 287:1.
    V tej nalogi primerjamo tri pristope na naboru 32.141 prehodov vozil:
    (1) klasični filter Savitzky-Golay z iskanjem vrhov, (2) 1D U-Net konvolucijsko
    nevronsko omrežje (CNN, 30\,M parametrov) in (3) temporalno konvolucijsko
    omrežje (TCN, 2{,}7\,M parametrov) z devetimi dilatiranimi rezidualnimi bloki,
    katerih receptivno polje znaša 2.045 vzorcev. Oba modela globokega učenja sta
    naučena s tehtano binarno navzkrižno entropijo (pos\_weight\,=\,287).
    Najboljši model (TCN) dosega F1\,=\,0{,}998, natančnost\,=\,0{,}997,
    priklic\,$\approx$\,1{,}000 in povprečno absolutno napako časovne lokacije
    0{,}24 vzorca, pri čemer prekaša CNN kljub 11-kratnemu zmanjšanju parametrov.
}
{
    Bridge Weigh-In-Motion (B-WIM) systems estimate the gross vehicle weight of
    heavy goods vehicles by analysing the dynamic strain signal recorded as a
    vehicle crosses a bridge. A critical preprocessing step is accurate axle
    detection --- locating the exact sample positions of each axle passage in a
    1,300-sample signal. This is a challenging binary sequence-labelling problem
    due to severe class imbalance: only 0.35\,\% of samples are positive (axle
    events), yielding a 287:1 negative-to-positive ratio. This thesis compares
    three approaches on a dataset of 32,141 vehicle crossings: (1) a classical
    Savitzky-Golay filter and peak-detection baseline, (2) a 1D U-Net
    Convolutional Neural Network (CNN, 30\,M parameters), and (3) a Temporal
    Convolutional Network (TCN, 2.7\,M parameters) with nine dilated residual
    blocks and a receptive field of 2,045 samples. Both deep learning models are
    trained with a weighted binary cross-entropy loss (pos\_weight\,=\,287).
    The TCN achieves F1\,=\,0.998, Precision\,=\,0.997, Recall\,$\approx$\,1.000,
    and a Mean Absolute Timing Error of 0.24 samples on the validation set,
    outperforming both the CNN and the classical baseline. The TCN achieves this
    with 11 times fewer parameters than the CNN, demonstrating that architecture
    design outweighs model scale for B-WIM axle detection.
}

% ---------------------------------------------------------------------------
% Acknowledgments
% ---------------------------------------------------------------------------
\acknowledgments
{
    % TODO: Write your acknowledgments here.
    I would like to thank my supervisor for their guidance and support throughout
    this project.
}

% License page (English version).
\input{license.tex}

% ---------------------------------------------------------------------------
% List of abbreviations
% ---------------------------------------------------------------------------
\abbreviation{B-WIM}{Bridge Weigh-In-Motion}
\abbreviation{BCE}{Binary Cross-Entropy loss}
\abbreviation{CNN}{Convolutional Neural Network}
\abbreviation{F1}{F1-score (harmonic mean of precision and recall)}
\abbreviation{GVW}{Gross Vehicle Weight}
\abbreviation{HGV}{Heavy Goods Vehicle}
\abbreviation{MATE}{Mean Absolute Timing Error}
\abbreviation{ReLU}{Rectified Linear Unit}
\abbreviation{TCN}{Temporal Convolutional Network}

\begin{document}
    \input{uvod.tex}        % Chapter 1 — Introduction
    \input{background.tex}  % Chapter 2 — Background and Related Work
    \input{metodologija.tex}% Chapter 3 — Methodology
    \input{results.tex}     % Chapter 4 — Results and Discussion
    \input{zakljucek.tex}   % Chapter 5 — Conclusion

    \backmatter
    % English theses must include an extended Slovene abstract (4,000–10,000
    % characters incl. spaces) as the last numbered chapter before References.
    % Uncomment and fill in abstract_sl.tex when ready.
    % \input{abstract_sl.tex}

    \input{priloga.tex}     % Appendix A — (optional)
\end{document}
